% -*- TeX-master: "ijcai18.tex" -*-

\section{Counterfactuals and prevention}\label{sec:inhibition}

\ifshort

The diagram shown Figure~\ref{fig:cex} extends the one in
Figure~\ref{fig:dumb-story} and explains the counterfactual dependency
between $pk$ and $p$ in \RefTrace{}. The dotted node corresponds to a
\emph{counterfactual event}, which is absent from \RefTrace{}. It is
related to \emph{factual} events by prevention arrows, shown in
red. These arrows can be read as follows: ``$pk$ prevents $u$, which
prevents $p$''. In this section, we give a rigorous account of
prevention by connecting events in a factual trace to events in a
cognate counterfactual trace as produced by counterfactual
resimulation, and vice versa.

\else

The diagram shown Figure~\ref{fig:cex} extends the one in
Figure~\ref{fig:dumb-story} and explains the counterfactual dependency
between $pk$ and $p$ in \RefTrace{}. The dotted node corresponds to a
\emph{counterfactual event}, which is absent from \RefTrace{}. It is
related to \emph{factual} events by prevention arrows, shown in
red. These arrows can be read as follows: ``$pk$ prevents $u$, which
prevents $p$''. In this section, we give a precise semantics to such
diagrams and discuss how they can be generated systematically.

As discussed in section~\ref{subsec:dumb-story}, a causal narrative
like the one in Figure~\ref{fig:dumb-story} is generated from a trace,
which usually results from a compression process, and whose events are
organized in a directed acyclic graphs, edges being enablement arrows.
Enablement arrows are easy to define in the context of a trace, and we
give a formal definition in section~\ref{subsec:inhibition}.
Prevention arrows are trickier because they must relate events that
happened to events that did not. For this reason, previous
work~\cite{DanosEtAl-CONCUR07} fails to provide a proper account of
them.  Our insight is to use counterfactual traces to give a rigorous
account of prevention using arrows that connect events from a factual
trace to events in an associated counterfactual trace and vice versa.

\fi

\ifshort \vspace{-0.45cm} \else \vspace{-0.3cm} \fi

\paragraph{Remark on notation} Although we used the letter $e$ to
denote potential events in previous sections, we use it to denote
\emph{events} in this section. We write $\TIME{e}$ the time of
occurence of event $e$. Finally, an event $e$ occuring at time $t$ is
said to be executable in trace $\tau$ if the associated potential
event is executable in mixture $\TSTATE{t}{\tau}$.


\subsection{Prevention in counterfactual experiments}
\label{subsec:cex}\label{subsec:inhibition}

A \textit{counterfactual experiment} is a triple
$(\tau, \iota, \tau')$ for which there exists a schedule $\sigma$ such
that $\tau = T(\sigma)$ and $\tau' = \ATRAJ{}(\sigma)$. Such triples
are produced by counterfactual resimulation.

To formalize enablement and prevention we need to define some terms
used in Kappa to talk about events. Without loss of generality, we
shall assume that a site on an agent carry either a binding state or
a tagging state but not both. The
\textit{value} of a tagged site in a mixture is its current tag and
the value of a binding site is either $\textsc{free}$ or
$\textsc{bound-to}(s)$, where $s$ identifies another site in the
mixture.  An event is said to \emph{modify} a site if it changes its
value.  An event is said to \emph{test} a site $s$ for value $v$ if
the left-hand side of the associated rule requires $s$ to have value
$v$ for the event to be executable. For example, event $p$ in
\RefTrace{} tests three sites and modifies one.

\input{figures/cex}

\begin{definition}[Enablement]
  Let $\tau$ be a trace and $e, e'$ two events in $\tau$.  We say that
  $e$ enables $e'$ if $e$ is the last event before $e'$ that modifies
  some site to the value it is tested for by $e'$.
\end{definition}

\begin{definition}[Prevention]
  Let $(\tau, \iota, \tau')$ be a counterfactual experiment. An event
  $e$ that occurs at time $t$ in $\tau$ is said to prevent an event
  $e'$ that occurs at time $t'$ in $\tau'$ if all of the following
  hold:
  \begin{inparaenum}[(1)]
  \item \label{inhibition:time} $t < t'$ ;
  \item \label{inhibition:breaks} there exists a site $s$ such that
    $e$ is the last event in $\tau$ before $t'$ that modifies the
    value of $s$ away from what $e'$ tests it for ;
  \item \label{inhibition:nointf} there are no events in $\tau'$ that
    modify $s$ during the time interval $(t, t')$.
  \end{inparaenum}
  The same definition holds switching $\tau$ and $\tau'$.
\end{definition}

Counterfactual experiments can be represented as directed acyclic
graphs like the one in Figure~\ref{fig:cex}. Such a graph features
three kinds of nodes:
\begin{inparaenum}[]
\item events that are proper to the factual trace (thick solid nodes),
\item events that are proper to the counterfactual trace (dotted
  nodes) and
\item events that are common to both traces (thin solid nodes).
\end{inparaenum}
% Edges consist in enablement and prevention arrows.

As illustrated Figure~\ref{fig:cex}, the influence of $pk$ on $p$ in
our example is mediated by the counterfactual event $u$. Such
mediating events always exist, as stated by the following theorem.

\begin{theorem}[Completeness of enablement and prevention]
  \label{thm:completeness}
  Let $(\tau, \iota, \tau')$ be a counterfactual experiment and $e$ an
  event that belongs only to $\tau$. Then, there exists an event
  $\hat e \in \tau$ that is blocked by $\iota$ and there is a directed
  path from $\hat e$ to $e$ with an even number of prevention arrows.
\end{theorem}
\noindent This theorem states that counterfactual dependencies can
always be explained in terms of enablement and prevention relations
between individual events. A proof is in
Appendix~\ref{ap:completeness}. This result establishes a bridge
between two different visions of causality: the vision based on event
structures \cite{winskel1986event} that is dominant in the concurrency
community and the vision based on counterfactuals which is dominant in
the causal inference community \cite{pearl2009causality}.

\subsection{Compression of counterfactual experiments}

Counterfactual experiments produced with counterfactual resimulation
are usually very large. They typically feature a lot of redundancy,
including events that are irrelevant to the outcome of interest or
futile cycles as discussed in section~\ref{subsec:dumb-story}. A
compression step is usually necessary before concise causal narratives
can be extracted from such experiments. However, the two traces of a
counterfactual experiment cannot be compressed separately following
the procedure described in section~\ref{subsec:dumb-story}, as there
is no guarantee that the compressed traces can still be generated from
a unique schedule to form a valid counterfactual experiment. Instead,
compressing a counterfactual experiment consists in extracting a
minimal \emph{valid} sub-experiment.

A counterfactual experiment $(\tau_1, \iota, \tau_1')$ is said to be a
\emph{sub-experiment} of $(\tau_2, \iota, \tau_2')$ if $\tau_1$ is a
sub-trace of $\tau_2$ and $\tau_1'$ is a sub-trace of $\tau_2'$. Also,
valid counterfactual experiments can be characterized as follows.

\begin{proposition}%[Characterization of counterfactual experiments]
  \label{prop:valid-cex}
  A triple $(\tau, \iota, \tau')$ is a valid counterfactual experiment
  if and only if all of the following hold:
  \begin{inparaenum}[(1)]
  \item \label{valid-cex:valid-traces} both $\tau$ and $\tau'$ are
    valid traces ;
  \item \label{valid-cex:no-blocking} no event of $\tau'$ is blocked
    by $\iota$ ;
  \item \label{valid-cex:co-occur} for every event $e \in \tau$ such
    that $e \notin \tau'$, then either $e$ is not executable in
    $\tau'$ or $e$ is blocked by $\iota$ ;
  \item \label{valid-cex:co-occur2} for every event $e' \in \tau'$
    such that $e' \notin \tau$, then $e'$ is not executable in $\tau$.
  \end{inparaenum}
\end{proposition}

\noindent Compressing a counterfactual experiment consists in finding
a minimal valid sub-experiment such that
\begin{inparaenum}[(i)]
\item the outcome of interest appears in the factual trace but not in
  the counterfactual trace and
\item events that are blocked by $\iota$, or on which the outcome of
  interest was shown to be counterfactually dependent in previous
  analyses, are kept in the factual trace.
\end{inparaenum}
Because these constraints along with the properties featured in
Proposition~\ref{prop:valid-cex} can be encoded as boolean
satisfiability constraints, compressing a counterfactual experiment
can be done using standard \textsc{sat}-solving techniques.

