\section{Practical use of  counterfactual resimulation}
\label{sec:cosim-practice}

Returning to our example, let's see how we can use counterfactual
resimulation to improve our understanding of the typical causal
narrative leading to the phosphorylation of a substrate. First, how
may one come to suspect that the narrative that is shown
Figure~\ref{fig:dumb-story} is incomplete \textit{a priori} ? And
then, how should one know what interventions might be worth trying in
order to discover some additional causal structure through
counterfactual resimulation?

There is no perfect answer here but there are heuristics we can
use. In our example, one may notice by running repeated simulations
that when rule $p$ triggers, the kinase involved is very likely to be
phosphorylated. The fact that the causal narrative
Figure~\ref{fig:dumb-story} does not provide any insight into this
phenomenon is a sign that we should investigate whether it is
an accidental correlation or revealing of an unknown causal effect. In
order to do so, it is natural to intervene by blocking the last event
phosphorylating the kinase involved in $p$ in sampled traces. More
generally, when
\begin{inparaenum}[(i)]
\item an event $e$ in a candidate causal narrative is executed
  unusually often in a context which is more specific than what is
  required by the left-hand side of the associated rule and
\item this effect cannot be explained by the action of other events in
  the same causal narrative,
\end{inparaenum}
\textit{then} exploring counterfactual traces in which we block the last event
responsible for at least a part of the unexplained context in the
triggering a $e$ is likely to be worthwhile.

In our example, this heuristic suggests that we should block the last
event responsible for the phosphorylation of the kinase before the
substrate gets phosphorylated itself in sampled traces. For the
particular case of \RefTrace{} of
section~\ref{sec:example}, we ask: how likely is it that ``$p$ would
have happened had $pk$ not happened'' ?  This can be translated
literaly in the question of evaluating $\mathbf{P}(\CFST{})$ where
\begin{inparaenum}[(i)]
  % \item $\tau$ is trace~(\ref{example-trace})
\item $\BLOCKED{\iota}{e'} = (e = pk)$ and
\item $\psi[\tau'] = (p \in \tau')$.
\end{inparaenum}
This choice of $\iota$ and $\psi$ is reasonable most of the
time but some variations may be prefered occasionally. For example, if
the rule $pk$ for phosphorylating is very fast, blocking only a single
instance of the potential event associated with $pk$ may not have a
very large effect as a new instance would trigger almost immediately
after in most counterfactual traces, cancelling the effect of the
intervention.  In this case, it is useful to make $\iota$ block every
instance of the potential event associated with $pk$ in a defined
timeframe.